---
title: "Seasonal Arima(1,1,1)x(1,1,1),12"
output: html_notebook
---

Fitting an ARIMA(1,1,1)x(1,1,1)_12
January 25, 2018

Predict the MONTHLY US residential energy consumption for the next 2 years for the 'Short-Term Energy Outlook'

In my last post I went through a non-seasonal ARIMA, looking at annual values, but what if the values I am looking for are not annual, but monthly? 

  No problem! Just change the y limits and gather monthly datapoints. Pretty simple, right?

Well, not really.. 

Whether this is how nature works or just the behavior of human beings, monthly values tend to rise and fall according to season. Winter Holidays seem to bring in the highest volume of consumption during a period. Ice cream sales on aggregate decrease significantly during the winter (though the amount I eat seemed rather consistent growing up).

Likewise, energy consumption in the US functions in a similar way. Let's take a look at how we can forecast the 2-year energy consumption for US residents. Data comes from http://eia.gov which shows "energy information" including energy sources and news events that may adversely or conversely affect energy acquisition in the US. Finding the right type of data gets a little hairy, but luckily there are filters that let you adjust parameters before you download any of the data. I decided to just scrape all the data and filter it myself. We are going to rip the energy data from the 'short-term energy outlook' found in the read.csv link. 
```{r}
data1 <- read.csv("http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T02.01")
#subset to TERCBUS Total Energy Consumed by the Residential Sector
data2 <- subset(data1,MSN=="TERCBUS")

#subset to "my lifetime" because energy consumption 
#has changed drastically from the past until now.
data3 <- subset(data2,data2$YYYYMM>199100)

#remove yearly total (coded "month 13", every 13th obs. We don't want that..)
data4 <- subset(data3,data3$YYYYMM%%100 != 13)
energy <- as.numeric(data4$Value)

rm(data1)             # delete the original data set ... don't need the extra information.
```
This is what the data will now look like in head() (above) and the plot (below).

```{r}
plot(energy, 
     type = "b",
     main = "Energy Consumption",
     xlab = "Month",
     ylab = "consumption (trillion Btu)"
     )
```
Well, how odd! the values seem like they move in a seasonal pattern with a periodical movement. I wonder why? Most likely due to seasonal consumption of energy. The two peaks might be explained by the energy used to heat homes in the winter and air conditioning during the summer. Otherwise, the plot doesn't seem to have any abnormalities.

Next, we can do a similar process in identifying the parameter estimate and errors. But wait.. What are all these ones doing and what does this twelve mean? Well, if you have time, you might want to take a look at the explanations under the arima(1,1,1). The next (1,1,1)_12 stands for the seasonal component of the arima model.The first three ones perform an autoregression of additive data corresponding year to year while the last three ones take into consideration the oscillating month to month pattern (hence the 12 for the twelve months in a year). Of course, these values can all be adjusted for further fine-tuned investigation yet this model seems to suffice for constant patterns. 

```{r}
#install.packages("asta")                               #Don't forget to install necessary packages!!
#library("astsa", lib.loc="~/Library/R/3.4/library")


#Fit an ARIMA(1,1,1)X(1,1,1),12 model and report the parameter estimates and standard errors.
energy.out <- sarima(energy,1,1,1,1,1,1,12)

#table of estimates
energy.out$ttable
```

Getting back to the table of estimates we notice that there is an ar1, ma1 corresponding to phi and epsilon of the annual data. There is also sar1, sma1 which are tied to phi and epsilon of the seasonal data. Where did the constant mu go? Well, this gets rather computationally greedy, but it can be simplified. Since arima estimates by maximum likelihood, mu gets subtracted away from seasonal differences as it intersects with the annual information.

Now back to the fun portion! We use the same sarima.for function and put the same 7 values, but also specify 27 units ahead. Because we only had data from september 2017, we wanted to account for the last three months that we missed. Of course, you are free to play around with these parameters.

```{r}
#Forecast the US resindential energy consumption for the next 2 years.
#Next to years 2918-2919
energy.future <- sarima.for(energy, n.ahead = 27, 1,1,1,1,1,1,12)
energy.future$pred


```
Now look at that! We have a simple graphic forcasting the next 27 months and the individual values. This was definitely less coding than a non-seasonal arima despite how fearful the data looks. There was no need to transform the data, and half the lines of code. 

One downfall to the sarima graphic is that we cannot specify the limits, title, or labels. So, without any explanation, I will attempt to recreate the same graphic above, below.
```{r}
#Report a table of predictions and 95% predictions intervals.
L <- energy.future$pred - 2 * energy.future$se
U <- energy.future$pred + 2 * energy.future$se
cbind("prediction"=energy.future$pred,"lower"=L,"upper"=U)

#A publication quality graphic showing the historical values and the 2 year predictions
#(with uncertainty) of US residential energy consumption
plot(energy,
     type="o",
     ylab="consumption (trillion Btu)",
     xlab = "month",
     xlim = c(250,350),
     main = "Total Energy Consumed by the Residential Sector")
abline(v=321, col = "dark blue")
length(322:348)
lines(322:348,L,col="blue",lty=2)                                         #lower bounds
lines(322:348,U,col="blue",lty=2)                                         #upper bounds
polygon(c(322:348,rev(322:348)),c(L,rev(U)),col = "light gray", border=NA)#Filling in the lines between the upper
lines(322:348,energy.future$pred,col="red",type="b",pch=8)                #and lower bounds in the prediction.
```


I tried! Not sure if I like the aesthetics as much as the other one, but close enough right?

The seasonal arima model is impressive as it flexibily represents several varieties of time series with simplicity. Of course, there are always weaknesses. For example, the same assumptions apply with the non-seasonal arima. It does not take into account unsuspected predictor variables or time deterministic trends. It looks from year to year. Additionally, the weakness with this type of model is that the addition of more data does not necessarily mean more data points. It means more detailed data points that are included in the time frame. If more data is necessary, we will need quality descriptive data points in between each seasonal trend. If you have time on your own, give it a try!

One place you can look to practice with certain types of data would be qlik.com. Qlick has an enormous databse of information available to the public. All of the data is stored by an internal server that was previously owned by datamarket.com. Take a look at some of the data below and try it on your own.

ie. Monthly sales of Tasty Cola
#https://datamarket.com/data/set/22xx/monthly-sales-of-tasty-cola#!ds=22xx&display=line







